---
title: "Covid 19 Data Analysis"
author: "Seif Elkateb"
date: "2025-03-04"
output: html_document
---

The Covid-19 pandemic in 2020 was unprecedented on the 21th century and there has been millions of global cases and also deaths all over the world. By analyzing the  data we then could then learn how to better take actions in the future if another pandemic in the future. 

### Goals of the analysis

- Track the progression of the pandemic 
- Identify the hot spots and areas that needs better resources 
- Evaluate the effectiveness of public heath measures 
- Evaluate the effectiveness of the polices set by the government 


 The data used in this analysis can be accessed from this  [COVID-19 Data Repository.](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)


### Libraries that will be used 

- Tidyverse
- Lubridate

```{r Libraries_Setup, message=FALSE,include=TRUE,warning=FALSE}
#using the tidyverse library
library(tidyverse)
library(lubridate)
```

### Importing the data 

```{r importing_data , message=FALSE}

#preparing the data and importing them 

us_cases<-read_csv("./Data/time_series_covid19_confirmed_US.csv")
global_cases<-read_csv("./Data/time_series_covid19_confirmed_global.csv")
us_deaths<-read_csv("./Data/time_series_covid19_deaths_US.csv")
global_deaths<-read_csv("./Data/time_series_covid19_deaths_global.csv")
look_up<-read_csv("./Data/UID_ISO_FIPS_LookUp_Table.csv")

```
### Tidying and transforing the data 

#### **There are 4 csv files** 

- **Us Confirmed Cases** 
- **Us Death Cases** 
- **Global Confirmed Cases** 
- **Global Deaths** 


### Global Data Tidying

after looking at the global cases and deaths csv files (**global_cases,global_deaths**). I need to do the following. 

- The Lat and Long columns are not needed for my analysis on the covid 19 data

- Each date has its own column and that needs to be modified by creating the following columns:
  - Date column where each observation(row) will have one of the dates originally used as columns 
  - Cases which will contain the cases on that specific date for the specific country/region 
  - Deaths in the case of global_deaths.csv file and it will contain death count for specific country/region
  - The Date on all the csv files is not Date type it's character type and this needs to be changed to the correct type


```{r Tidying_Global_Data,include=TRUE, echo=TRUE}
# There are 2 csv files related to global cases  : global cases and global deaths 

#selecting only the Province/state , country/region , besides creating date and cases in global_cases

global_cases<- global_cases %>% select(-c('Lat','Long'))%>%pivot_longer(cols=-c('Province/State','Country/Region'),names_to="date",values_to ="cases")%>% mutate(date=mdy(date))
                                                                        
#selecting only the Province/state , country/region , besides creating date and deaths in global_deaths

global_deaths<- global_deaths %>% select(-c('Lat','Long'))%>%pivot_longer(cols=-c('Province/State','Country/Region'),names_to="date",values_to ="deaths")%>% mutate(date=mdy(date))

```

now i will create object called global which will contain the data from both the global_deaths and global_cases after tidying them . Also i will rename province/State and Country/region columns for a better readability 

```{r Global_Data_tidying_Extra,include=TRUE,echo=TRUE,message=FALSE}

global<- global_cases%>% full_join(global_deaths) %>% rename(Country_Region='Country/Region',Province_State='Province/State')

```

### Summary of the data 

**There are some observations such as**

- Some dates on the dataset have **0 cases and deaths** and i need to decide whether i want to include them on my analysis or not 
- Last Date on the dataset is **2023-03-09**
- Max Number of cases is **103802702**
- Max Number of deaths is **1123836**

```{r summary_of_data, echo=TRUE,include=TRUE}
summary(global)
```


### Decision After analyzing the Global data

- After reviewing the summary of the data i decided i need to filter the observations and only include the ones with cases >0
- now if i do that i will be able to analyze the data better and i will be able how the death column is correlated to the cases column
- some dates have a lot of cases but zero deaths!! 


```{r filtering_global_data, include=TRUE,echo=TRUE}
global <- global %>% filter(cases>0)

summary(global)
```



### US Data Tidying

- no need for lat, long columns
- dates have columns and need to be transformed into a column called date and corresponding deaths or cases 
- dates need to be transformed to date obj
- province_state and country_region don't need to be renamed 
- UID, ISO2 , ISO3 , CODE 3 ,Fips 
- Admin 2 , combined key ,population need to be present 
- The US_deaths have population column while US_cases don't have that column



```{r us_data_tidying, include=TRUE,echo=TRUE}
# There are 2 csv files related to US   : global cases and global deaths 

#selecting only the Province/state , country/region , besides creating date and cases in global_cases

us_cases<- us_cases %>% select(-c('UID','Lat','Long_','iso2','iso3','code3','FIPS'))%>%pivot_longer(cols=-c('Province_State','Country_Region','Admin2','Combined_Key'),names_to="date",values_to ="cases")%>% mutate(date=mdy(date))



us_deaths<- us_deaths %>% select(-c('UID','Lat','Long_','iso2','iso3','code3','FIPS'))%>%pivot_longer(cols=-c('Province_State','Country_Region','Admin2','Combined_Key','Population'),names_to="date",values_to ="deaths")%>% mutate(date=mdy(date))
```


### Us Data 

```{r us_dataset,include=TRUE,echo=TRUE,message=FALSE,warning=FALSE}
us<-us_cases %>%full_join(us_deaths)

summary(us)
```

### summary US data

```{r summary_us_data,include=TRUE,echo=TRUE}

#us dataset sumamry after filtering 
summary(us)
```

### Final Transformation of the data 

- Global data does not have the population column
- Global data does not have the combined key and i would either have to add it to the global data or remove it 

```{r final_data_transformation,echo=TRUE,include=TRUE}
global<- global %>% unite('Combined_Key',c(Province_State,Country_Region),sep=", ",na.rm=TRUE,remove=FALSE)



look_up<-look_up%>%select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global<- global %>% left_join(look_up,by=c('Province_State','Country_Region'))%>% select(-c('UID','FIPS'))
summary(global)

```


### Data Visualization


- Now moving for the data visualization but before that we need to do more transformation to the data frame and maybe create new data frames that we can do visualizations on them 


```{r data_transformation_extraa,echo=TRUE,include=TRUE,message=FALSE,warning=FALSE}


# the sum of the population in US 
# this is the sum of the population of the US




US_by_state <- us %>%
group_by(Province_State, Country_Region, date) %>%
summarize(cases = sum(cases), deaths = sum(deaths),
Population = sum(Population)) %>%
mutate(deaths_per_mill = deaths *1000000 / Population) %>%
select(Province_State, Country_Region, date,
cases, deaths, deaths_per_mill, Population) %>%
ungroup()



US_totals <- us %>%
group_by(Country_Region, date) %>%
summarize(cases = sum(cases), deaths = sum(deaths),
Population = sum(Population)) %>%
mutate(deaths_per_mill = deaths *1000000 / Population) %>%
select(Country_Region, date,
cases, deaths, deaths_per_mill, Population) %>%
ungroup()



tail(US_totals)

```

### Data Visualization 

**Cases Versus Deaths in US all country **


```{r data_visualization, echo=FALSE}


# what is the Max Date on the Data [1] "2023-03-09" which is 3 years since the covid pandemic 

max(US_totals$date)

# what is the Max Deaths on the Data [1] "1123836" 

max(US_totals$deaths)

US_totals %>%
filter(cases > 0) %>%
ggplot(aes(x = date, y = cases)) +
geom_line(aes(color = "cases")) +
geom_point(aes(color = "cases")) +
geom_line(aes(y = deaths, color = "deaths")) +
geom_point(aes(y = deaths, color = "deaths")) +
scale_y_log10() +
theme(legend.position="bottom",
axis.text.x = element_text(angle = 90)) +
labs(title = "COVID19 in US", y= NULL)







```
Cases Versus Deaths in specific States


```{r data_visualization_state,echo=TRUE,include=TRUE,warning=FALSE,message=FALSE}



state <- "New York"
US_by_state %>%
filter(Province_State == state) %>%
filter(cases > 0) %>%
ggplot(aes(x = date, y = cases)) +
geom_line(aes(color = "cases")) +
geom_point(aes(color = "cases")) +
geom_line(aes(y = deaths, color = "deaths")) +
geom_point(aes(y = deaths, color = "deaths")) +
scale_y_log10() +
theme(legend.position="bottom",
axis.text.x = element_text(angle = 90)) +
labs(title = str_c("COVID19 in ", state), y= NULL)



## covid 19 in Alabama

state <- "Alabama"
US_by_state %>%
filter(Province_State == state) %>%
filter(cases > 0) %>%
ggplot(aes(x = date, y = cases)) +
geom_line(aes(color = "cases")) +
geom_point(aes(color = "cases")) +
geom_line(aes(y = deaths, color = "deaths")) +
geom_point(aes(y = deaths, color = "deaths")) +
scale_y_log10() +
theme(legend.position="bottom",
axis.text.x = element_text(angle = 90)) +
labs(title = str_c("COVID19 in ", state), y= NULL)

```
### Daily Cases Over Time for Global Data 


```{r daily_global,include=TRUE,echo=TRUE,message=FALSE,warning=FALSE}

# Summarize daily deaths globally
global_daily_deaths <- global_deaths %>%
  group_by(date) %>%
  summarize(daily_deaths = sum(deaths)) %>%
  ungroup()

# Create line plot
ggplot(global_daily_deaths, aes(x = date, y = daily_deaths)) +
  geom_line(color = "red") +
  theme_minimal() +
  labs(title = "Daily COVID-19 Deaths Globally", x = "Date", y = "Daily Deaths")

```
### Summarize total cases by country and select top 10

```{r bar_plot,include=TRUE,echo=TRUE,message=FALSE,warning=FALSE}

# Summarize total cases by country and select top 10
global_cases_by_country <- global_cases %>%
  group_by(`Country/Region`) %>%
  summarize(total_cases = sum(cases)) %>%
  ungroup() %>%
  top_n(10, total_cases)

# Create bar plot
ggplot(global_cases_by_country, aes(x = reorder(`Country/Region`, -total_cases), y = total_cases)) +
  geom_bar(stat = "identity", fill = "blue") +
  theme_minimal() +
  labs(title = "Top 10 Countries by Total COVID-19 Cases", x = "Country", y = "Total Cases") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


### Modeling Data 

```{r modeling_global,include=TRUE,echo=TRUE,message=FALSE,warning=FALSE}

US_state_totals <- US_by_state %>%
group_by(Province_State) %>%
summarize(deaths = max(deaths), cases = max(cases),
population = max(Population),
cases_per_thou = 1000* cases / population,
deaths_per_thou = 1000 * deaths / population) %>%
filter(cases > 200000, population > 0)


mod <-lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)


summary(mod)

US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))



US_tot_w_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") + geom_point(aes(x = cases_per_thou, y = pred), color = "red")






```





the model does a reasonably good job of predicting at alot of points in the data but's it needs alot of modification and maybe adding additional factors to the model to optimize the calculations and predicted deaths per thousands. 



###Sessio Info


```{r session_info}
sessionInfo()
```


### Conclusion And Bias Identification

This analysis aimed to track the progression of the COVID-19 pandemic as of 2025, identify hotspots, evaluate public health measures, and assess government policies using data from the COVID-19 Data Repository. Employing the (tidyverse and lubridate libraries), i wrangled and transformed global and US-specific data frames, converting date to the right format, pivoting data  and searching for any inconsisntencies in the data.Im not from US and my country had different measures to deal with the covid pandemic , also the public health system is very different . So to mitigate any bias i tried to use all the variables possible and do my analysis multiple times after analyzing the visualizations and models 












